An antibody-escape estimator for mutations to the SARS-CoV-2 receptor-binding domain

### 2022-03-06 paper/paper.tex
change "calculator" -> "estimator"

\title{An antibody-escape calculator for mutations to the SARS-CoV-2 receptor-binding domain}

@@

\begin{abstract}
A key goal of SARS-CoV-2 surveillance is to rapidly identify viral variants with mutations that reduce neutralization by polyclonal antibodies elicited by vaccination or infection.
Unfortunately, direct experimental characterization of new viral variants lags their sequence-based identification. 
Here we help address this challenge by aggregating deep mutational scanning data into an ``escape calculator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
The calculator can be used to intuitively visualize how mutations impact polyclonal antibody recognition, and score the expected antigenic effect of combinations of mutations.
These scores correlate with neutralization assays performed on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
An interactive version of the calculator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
\end{abstract}

@@

Here we take a step towards addressing this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high}.
The resulting ``escape calculator'' enables qualitative visualization and quantitative scoring of the antigenic effects of arbitrary combinations of mutations.
Importantly, the escape calculator is based on simple transformations of direct experimental measurements, and does not involve complex black-box computational methods.

@@

\subsection{Aggregating deep mutational scanning data for 33 human antibodies yields a realistic escape calculator}
The toy example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
To create an escape map for an antibody mix that more realistically represents actual human sera, we aggregated previously generated deep mutational scanning data for 33 neutralizing antibodies elicited by SARS-CoV-2.
These antibodies were isolated from a variety of patient cohorts within the first year of the pandemic (see Methods for details).
An assumption of the analysis that follows is that an equipotent mixture of these 33 antibodies represents the neutralizing activity of human sera; we emphasize that this assumption is imperfect since in reality the antibodies were chosen for prior study for a variety of ad hoc reasons.
The escape maps for all the individual antibodies can be interactively interrogated at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/}.

\begin{SCfigure*}[][]
\includegraphics[width=0.8\textwidth]{figures/escape_calc/escape_calc.pdf}
\caption{
An escape calculator generated by aggregating deep mutational scanning for 33 neutralizing antibodies targeting the SARS-CoV-2 RBD.
{\bf (A)} The blue line shows the extent of escape mediated by mutations at each site, as estimated by simply averaging the data for all the individual antibodies.
{\bf (B)} The blue line shows escape map after a mutation to site 484 (red point) ablates recognition by antibodies strongly targeting that site, while the gray line shows the original escape map in the absence of any mutations.
{\bf (C)} The escape map after mutating sites 417, 484, and 501 (the three RBD sites mutated in the Beta variant).
An interactive version of this figure is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}.}
\label{fig:escape_calc}
\end{SCfigure*}

@@

We recommend the reader explores the interactive escape calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, to perform calculations like those in Figure~\ref{fig:escape_calc} for arbitrary combinations of mutated RBD sites.
Such visual exploration of different combinations of mutations provides an intuitive sense of the antigenic structure of the RBD.

\subsection{The escape calculations correlate well with neutralization assays of human polyclonal sera against SARS-CoV-2 variants}
For each set of RBD mutations, we can define a quantitative score that represents the polyclonal antibody binding that remains after making these mutations.
This score is defined using the same principle as the site-wise escape calculator in the previous section: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain (see Methods for details).
This calculation is implemented in the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).

\begin{SCfigure*}[][]
\includegraphics[width=0.7\textwidth]{../results/neut_studies/neut_studies.png}
\caption{Correlation of calculated binding scores with experimentally measured fold changes in neutralization for SARS-CoV-2 variants and mutants.
The data of \citet{lucas2021impact} was generated using live virus and sera from vaccinated individuals who were (top) or were not (bottom) previously infected with SARS-CoV-2.
The data of \citet{uriu2021neutralization} and \citet{wang2021antibody} was generated using pseudovirus against convalescent (top) or vaccine (bottom) sera, with vaccine sera from Pfizer BNT162b2 or Moderna mRNA-1273 vaccines, respectively. 
The fold changes represent the geometric means over all subjects in each cohort.
An interactive version of this figure that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/neut_studies.html}.
}
\label{fig:neut_studies}
\end{SCfigure*}

To test how these escape-calculator scores compare to experimentally measured neutralization titers, we collated neutralization data from three previously published studies~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody}, each of which characterized sera from two patient cohorts against a variety of SARS-CoV-2 variants and mutants.
One can imagine many reasons why the escape-calculator scores might differ from the real neutralization titers: the calculator only considers RBD mutations, the antibodies used by the calculator might not accurately reflect the real mix in polyclonal sera, etc.
But despite all these potential caveats, the escape-calculator scores correlate quite well with the measured neutralization titers across all studies and cohorts (Figure~\ref{fig:neut_studies}).
Therefore, the simple and intuitive approach used by the calculator seems to accurately reflect the dominant features of polyclonal antibody escape in the RBD.

\subsection{The escape calculator suggests extensive antigenic change in the new Omicron variant}
We applied the escape calculator to the recently reported Omicron variant, which has 15 mutated sites in its RBD~\citep{ngs-sa2021update,deoliveira2021tweet}.
The calculated binding score for the Omicron variant is much lower than any other SARS-CoV-2 variants of concern, indicating extensive antibody escape (Figure~\ref{fig:Omicron}A).
The Omicron variant's calculated score is roughly equivalent to that of a polymutant spike (PMS20) that was artificially engineered in a pseudovirus by \citet{schmidt2021high} to maximize escape from polyclonal serum antibodies.
For comparison, \citet{schmidt2021high} measured that neutralization titers against this artificial PMS20 spike were reduced by $\sim$20- to $\sim$80-fold for sera from various cohorts of vaccinated and infected individuals.

@@

The site-level escape map for the Omicron variant's RBD is shown in Figure~\ref{fig:Omicron}B.
The Omicron RBD has lost most peaks of antibody binding relative to the original RBD.
Exploration of the mutations using the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} indicates that mutations at sites 484, 446, and 417 are the biggest drivers of this antigenic change, although other mutations also contribute.
The residual peaks in the map suggests the remaining antibody activity against the Omicron variant RBD could be further eroded by mutations at sites like 346, 378, 444, and 504.

\section{Discussion}
We have described an escape calculator that uses experimental data for 33 monoclonal antibodies to estimate the antigenic effects of arbitrary combinations of mutations to the SARS-CoV-2 RBD.
The key insight is to aggregate data for individual antibodies to define both which RBD sites are antigenically important, and which combinations of mutations have redundant versus synergistic effects on antigenic escape.
For instance, sites 417, 484, and 490 are all peaks of antibody escape.
But mutations at 484 and 490 have redundant effects since they generally escape the same antibodies, whereas mutations at 484 and 417 have synergistic effects since they generally escape different antibodies.

Another key aspect of our approach is the interactive visual implementation of the escape calculator, so the user can interrogate the effects of mutations (or combinations of mutations) simply by clicking (or shift-clicking) on sites.
This interactivity provides an intuitive understanding of the antigenic structure of the RBD, and shows how the calculator is performing simple transformations directly on experimental data.
We encourage interactive use of the calculator so it acts as an aid to augment human interpretation, rather than a black-box computational method.
However, we note that in addition to the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}.

There are caveats that should be kept in mind when using the escape calculator.
First, it only assesses the antigenic effects of mutations, and not how they impact other properties such as viral transmissibility.
Second, the calculator only considers sites in the RBD, and ignores mutations to other regions of spike.
Third, the calculator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
Fourth, the calculator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
Finally, and in our minds most significantly, the calculator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).

Despite all these caveats, the escape calculator yields binding scores that correlate with experimentally measured neutralization titers.
In addition, the actual antigenic evolution of SARS-CoV-2 seems to follow the principles captured by the escape calculator: variants of concern generally have combinations of mutations calculated to have synergistic effects on antigenic escape (eg, 417 and 484) rather than combinations calculated to have redundant effects (eg, 484 and 490).
We suspect the calculator works well because the RBD is the dominant target of neutralizing activity~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high} and the human antibody-response to the early Wuhan-Hu-1 RBD shares broad commonalities across individuals~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.

However, the situation will become more complex over time.
Currently, most humans with antibodies to SARS-CoV-2 have been exposed to a RBD antigen that is identical or very similar to that of the early Wuhan-Hu-1 strain.
Therefore, antigenic studies can reasonably define mutations in reference to that RBD, since it is what the antibodies target.
But as humans are exposed to more diverged RBD variants, it will become difficult to determine what reference to use to define antigenic mutations, since different individuals will have antibodies targeting different RBDs.
Additionally, differing exposure histories can leave individuals with different antibody specificities~\citep{cobey2017immune}, a process that is already starting to occur for SARS-CoV-2~\citep{greaney2021sars}.
So in the future, it will be necessary to stratify the data used by the escape calculator by which RBD variant elicited the antibodies, and aggregate data for antibodies that reflect the sera in question.
For this reason, we expect to continue adding to the data used by the escape calculator, and emphasize that it will change over time from the version described here, although we provide stable links to the current version in the Methods below.

@@

\section{Methods}
\subsection{Code and data availability}
The most up-to-date code and data used to implement the escape calculator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps}, and the version described in this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/tree/bioRxiv_v1}.

The data used by the escape calculator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/processed_data/escape_calculator_data.csv}, and the version used for this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_calculator_data.csv}.

A downloadable HTML version of the escape calculator is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/raw/main/docs/_includes/escape_calc_chart.html}, with the version described in this paper at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/docs/_includes/escape_calc_chart.html}.

@@

\subsection{Deep mutational scanning data used by the calculator}
The experimental data used by the escape calculator are drawn from seven previously published deep mutational scanning studies~\citep{greaney2021complete,greaney2021mapping,starr2021prospective,starr2021complete,starr2021sars,dong2021genetic,tortorici2021broad} and one unpublished dataset available at \url{https://github.com/jbloomlab/SARS-CoV-2-RBD_MAP_COV2-2955}.
In total, these studies contain data for 36 monoclonal antibodies.
Three of these antibodies (CR3022, S304, and S309) were elicited by infection with SARS-CoV-1 and so are excluded from the datasets used for the calculations in this paper, although the calculator has an option (\texttt{eliciting\_virus}) that allows optional inclusion of these antibodies.
The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and were initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on sequences reported by Regeneron, Eli Lilly, or their collaborators~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
The full deep mutational scanning data for all these antibodies are interactively displayed at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} and available in raw form at \url{https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv} (see \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_data.csv} for a stable version of the raw data corresponding to that used in this paper).

The deep mutational scanning measures an escape fraction for each tolerated RBD mutation against each antibody, which represents an estimate of how completely that mutation escapes antibody binding~\citep{greaney2021complete}.
We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions at each site.
The results reported in this paper use the sums as the site-level metric, although the calculator has an option (\texttt{escape\_metric}) to use the mean instead.
We also normalize the site-level escape metrics for each antibody to account for different strengths of antibody selection in different experiments using the approach described in \citet{greaney2021comprehensive} and implemented in \url{https://jbloomlab.github.io/dmslogo/dmslogo.utils.html#dmslogo.utils.AxLimSetter} with \texttt{min\_upperlim=1} and \texttt{max\_from\_quantile=(0.5, 0.05)}: essentially this corresponds to scaling the site escape values for each antibody so that a value of one corresponds to the larger of the maximum escape at a site or 20 times the median value across sites.
The results reported in this paper use the normalized site-level metrics, although the calculator has an option (\texttt{escape\_values\_normalized}) to also use the non-normalized values.

\subsection{Calculation of the impact of mutations}
The escape calculator determines the impact of mutating sites by calculating how much each antibody is escaped by mutations at each site, and adjusting its contribution to the overall polyclonal mix accordingly.

@@

Let $\mathcal{M}$ be the set of sites that are mutated.
Then for each antibody we compute the binding retained as $b_a\left(\mathcal{M}\right) = \left(\prod\limits_{r \in \mathcal{M}} \frac{\max_{r'}\left(x_{a,r'}\right) - x_{a,r}}{\max_{r'}\left(x_{a,r'}\right)}\right)^s$. 
Essentially, this equation means that if the RBD is mutated at a strong site of escape for an antibody $a$, much if the binding is lost (if it is the strongest site of escape, all binding is lost).
The variable $s$ represents how dramatically binding is lost for mutations at sites of escape that are not the strongest: larger values of $s$ means mutations at even moderate sites of escape reduce binding a lot.
In this paper we report calculations with $s = 2$, although the calculator has an option (\texttt{mutation\_escape\_strength}) to choose other values.
We then define the escape map after the mutations $\mathcal{M}$ as $\frac{1}{A}\sum_a x_{a,r} b_a\left(\mathcal{M}\right)$.
The calculator shows the escape map with no mutations in gray, and that after mutations in blue.

@@

\subsection{Implementation of interactive calculator}
The interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} is imlemented using Altair (\url{https://altair-viz.github.io/})~\citep{VanderPlas2018}, which is in turn built upon Vega-Lite~\citep{Satyanarayan2017}.

\subsection{Python module with batch-mode calculator}
A Python module that implements the calculations is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}, and has all the same options as the interactive calculator.

>>>>

\title{An antibody-escape estimator for mutations to the SARS-CoV-2 receptor-binding domain}

@@

\begin{abstract}
A key goal of SARS-CoV-2 surveillance is to rapidly identify viral variants with mutations that reduce neutralization by polyclonal antibodies elicited by vaccination or infection.
Unfortunately, direct experimental characterization of new viral variants lags their sequence-based identification. 
Here we help address this challenge by aggregating deep mutational scanning data into an ``escape estimator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
The estimator can be used to intuitively visualize how mutations impact polyclonal antibody recognition, and score the expected antigenic effect of combinations of mutations.
These scores correlate with neutralization assays performed on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
An interactive version of the estimator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
\end{abstract}

@@

Here we take a step towards addressing this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high}.
The resulting ``escape estimator'' enables qualitative visualization and quantitative scoring of the antigenic effects of arbitrary combinations of mutations.
Importantly, the escape estimator is based on simple transformations of direct experimental measurements, and does not involve complex black-box computational methods.

@@

\subsection{Aggregating deep mutational scanning data for 33 human antibodies yields a realistic escape estimator}
The toy example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
To create an escape map for an antibody mix that more realistically represents actual human sera, we aggregated previously generated deep mutational scanning data for 33 neutralizing antibodies elicited by SARS-CoV-2.
These antibodies were isolated from a variety of patient cohorts within the first year of the pandemic (see Methods for details).
An assumption of the analysis that follows is that an equipotent mixture of these 33 antibodies represents the neutralizing activity of human sera; we emphasize that this assumption is imperfect since in reality the antibodies were chosen for prior study for a variety of ad hoc reasons.
The escape maps for all the individual antibodies can be interactively interrogated at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/}.

\begin{SCfigure*}[][]
\includegraphics[width=0.8\textwidth]{figures/escape_calc/escape_calc.pdf}
\caption{
An escape estimator generated by aggregating deep mutational scanning for 33 neutralizing antibodies targeting the SARS-CoV-2 RBD.
{\bf (A)} The blue line shows the extent of escape mediated by mutations at each site, as estimated by simply averaging the data for all the individual antibodies.
{\bf (B)} The blue line shows escape map after a mutation to site 484 (red point) ablates recognition by antibodies strongly targeting that site, while the gray line shows the original escape map in the absence of any mutations.
{\bf (C)} The escape map after mutating sites 417, 484, and 501 (the three RBD sites mutated in the Beta variant).
An interactive version of this figure is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}.}
\label{fig:escape_calc}
\end{SCfigure*}

@@

We recommend the reader explores the interactive escape estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, to perform calculations like those in Figure~\ref{fig:escape_calc} for arbitrary combinations of mutated RBD sites.
Such visual exploration of different combinations of mutations provides an intuitive sense of the antigenic structure of the RBD.

\subsection{The escape calculations correlate well with neutralization assays of human polyclonal sera against SARS-CoV-2 variants}
For each set of mutated RBD sites, we can define a quantitative score that represents the polyclonal antibody binding that remains after mutating these sites.
This score is defined using the same principle as the site-wise escape estimator in the previous section: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain (see Methods for details).
This calculation is implemented in the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).

\begin{SCfigure*}[][]
\includegraphics[width=0.7\textwidth]{../results/neut_studies/neut_studies.png}
\caption{Correlation of calculated binding scores with experimentally measured fold changes in neutralization for SARS-CoV-2 variants and mutants (smaller values indicate worse neutralization).
The data of \citet{lucas2021impact} was generated using authentic SARS-CoV-2 and sera from vaccinated individuals who were (top) or were not (bottom) previously infected with SARS-CoV-2.
The data of \citet{uriu2021neutralization} and \citet{wang2021antibody} was generated using pseudovirus against convalescent (top) or vaccine (bottom) sera, with vaccine sera from Pfizer BNT162b2 or Moderna mRNA-1273 vaccines, respectively. 
The fold changes are geometric means over all subjects in each cohort.
An interactive version of this figure that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/neut_studies.html}.
}
\label{fig:neut_studies}
\end{SCfigure*}

To test how these escape-estimator scores compare to experimentally measured neutralization titers, we collated neutralization data from three previously published studies~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody}, each of which characterized sera from two patient cohorts against a variety of SARS-CoV-2 variants and mutants.
One can imagine many reasons why the escape-estimator scores might differ from the real neutralization titers: the estimator only considers RBD mutations, the antibodies used by the estimator might not accurately reflect the real mix in polyclonal sera, etc.
But despite all these potential caveats, the escape-estimator scores correlate quite well with the measured neutralization titers across all studies and cohorts (Figure~\ref{fig:neut_studies}).
Therefore, the simple and intuitive approach used by the estimator seems to accurately reflect the dominant features of polyclonal antibody escape in the RBD.

\subsection{The escape estimator suggests extensive antigenic change in the new Omicron variant}
We applied the escape estimator to the recently reported Omicron variant, which has 15 mutated sites in its RBD~\citep{ngs-sa2021update,deoliveira2021tweet}.
The calculated binding score for the Omicron variant is much lower than any other SARS-CoV-2 variants of concern, indicating extensive antibody escape (Figure~\ref{fig:Omicron}A).
The Omicron variant's calculated score is roughly equivalent to that of a polymutant spike (PMS20) that was artificially engineered in a pseudovirus by \citet{schmidt2021high} to maximize escape from polyclonal serum antibodies.
For comparison, \citet{schmidt2021high} measured that neutralization titers against this artificial PMS20 spike were reduced by $\sim$20- to $\sim$80-fold for sera from various cohorts of vaccinated and infected individuals.

@@

The site-level escape map for the Omicron variant's RBD is shown in Figure~\ref{fig:Omicron}B.
The Omicron RBD has lost most peaks of antibody binding relative to the original RBD.
Exploration of the mutations using the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} indicates that mutations at sites 484, 446, and 417 are the biggest drivers of this antigenic change, although other mutations also contribute.
The residual peaks in the map suggests the remaining antibody activity against the Omicron variant RBD could be further eroded by mutations at sites like 346, 378, 444, and 504.

\section{Discussion}
We have described an escape estimator that uses experimental data for 33 monoclonal antibodies to estimate the antigenic effects of arbitrary combinations of mutations to the SARS-CoV-2 RBD.
The key insight is to aggregate data for individual antibodies to define both which RBD sites are antigenically important, and which combinations of mutations have redundant versus additive effects on antibody binding.
For instance, sites 417, 484, and 490 are all peaks of antibody escape.
But mutations at 484 and 490 have redundant effects since they generally escape the same antibodies, whereas mutations at 484 and 417 have additive effects since they generally escape different antibodies.

Another key aspect of our approach is the interactive visual implementation of the escape estimator, so the user can interrogate the effects of mutations (or combinations of mutations) simply by clicking (or shift-clicking) on sites.
This interactivity provides an intuitive understanding of the antigenic structure of the RBD, and shows how the estimator is performing simple transformations directly on experimental data.
We encourage interactive use of the estimator so it acts as an aid to augment human interpretation, rather than a black-box computational method.
However, we note that in addition to the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}.

There are caveats that should be kept in mind when using the escape estimator.
First, the estimator only considers sites in the RBD, and ignores mutations to other regions of spike.
Second, the estimator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
Third, the estimator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
Finally, and in our minds most significantly, the estimator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).

Despite all these caveats, the escape estimator yields binding scores that correlate with experimentally measured neutralization titers.
In addition, the actual antigenic evolution of SARS-CoV-2 seems to follow the principles captured by the escape estimator: variants of concern generally have combinations of mutations calculated to have additive effects on antibody escape (e.g., 417 and 484) rather than combinations calculated to have redundant effects (e.g., 484 and 490).
We suspect the estimator works well because the RBD is the dominant target of neutralizing activity~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high} and the human antibody-response to the early Wuhan-Hu-1 RBD shares broad commonalities across individuals~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.

However, the situation will become more complex over time.
Currently, most humans with antibodies to SARS-CoV-2 have been exposed to a RBD antigen that is identical or very similar to that of the early Wuhan-Hu-1 strain.
Therefore, antigenic studies can reasonably define mutations in reference to that RBD, since it is what the antibodies target.
But as humans are exposed to more diverged RBD variants, it will become difficult to determine what reference to use to define antigenic mutations, since different individuals will have antibodies targeting different RBDs.
Additionally, differing exposure histories can leave individuals with different antibody specificities~\citep{cobey2017immune}, a process that is already starting to occur for SARS-CoV-2~\citep{greaney2021sars}.
So in the future, it will be necessary to stratify the data used by the escape estimator by which RBD variant elicited the antibodies, and aggregate data for antibodies that reflect the sera in question.
For this reason, we expect to continue adding to the data used by the escape estimator, and emphasize that it will change over time from the version described here, although we provide stable links to the current version in the Methods below.

@@

\section{Methods}
\subsection{Code and data availability}
The most up-to-date code and data used to implement the escape estimator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps}, and the version described in this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/tree/bioRxiv_v1}.

The data used by the escape estimator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/processed_data/escape_calculator_data.csv}, and the version used for this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_calculator_data.csv}.

A downloadable HTML version of the escape estimator is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/raw/main/docs/_includes/escape_calc_chart.html}, with the version described in this paper at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/docs/_includes/escape_calc_chart.html}.

@@

\subsection{Deep mutational scanning data used by the estimator}
The experimental data used by the escape estimator are drawn from seven previously published deep mutational scanning studies~\citep{greaney2021complete,greaney2021mapping,starr2021prospective,starr2021complete,starr2021sars,dong2021genetic,tortorici2021broad} and one unpublished dataset available at \url{https://github.com/jbloomlab/SARS-CoV-2-RBD_MAP_COV2-2955}.
In total, these studies contain data for 36 monoclonal antibodies.
Three of these antibodies (CR3022, S304, and S309) were elicited by infection with SARS-CoV-1 and so are excluded from the datasets used for the calculations in this paper, although the estimator has an option (\texttt{eliciting\_virus}) that allows optional inclusion of these antibodies.
The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and were initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on previously reported sequences~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
The full deep mutational scanning data for all these antibodies are interactively displayed at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} and available in raw form at \url{https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv} (see \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_data.csv} for a stable version of the raw data corresponding to that used in this paper).

The deep mutational scanning measures an escape fraction for each tolerated RBD mutation against each antibody, which represents an estimate of how completely that mutation escapes antibody binding~\citep{greaney2021complete}.
We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions across all tolerated mutations at each site.
The results reported in this paper use the sums as the site-level metric, although the estimator has an option (\texttt{escape\_metric}) to use the mean instead.
We also normalize the site-level escape metrics for each antibody to account for different strengths of antibody selection in different experiments using the approach described in \citet{greaney2021comprehensive} and implemented in \url{https://jbloomlab.github.io/dmslogo/dmslogo.utils.html#dmslogo.utils.AxLimSetter} with \texttt{min\_upperlim=1} and \texttt{max\_from\_quantile=(0.5, 0.05)}: essentially this corresponds to scaling the site escape values for each antibody so that a value of one corresponds to the larger of the maximum escape at a site or 20 times the median value across sites.
The results reported in this paper use the normalized site-level metrics, although the estimator has an option (\texttt{escape\_values\_normalized}) to also use the non-normalized values.

\subsection{Estimation of the impact of mutations}
The escape estimator determines the impact of mutating sites by calculating how much each antibody is escaped by mutations at each site, and adjusting its contribution to the overall polyclonal mix accordingly.

@@

Let $\mathcal{M}$ be the set of sites that are mutated.
Then for each antibody we compute the binding retained as $b_a\left(\mathcal{M}\right) = \left(\prod\limits_{r \in \mathcal{M}} \frac{\max_{r'}\left(x_{a,r'}\right) - x_{a,r}}{\max_{r'}\left(x_{a,r'}\right)}\right)^s$. 
Essentially, this equation means that if the RBD is mutated at a strong site of escape for an antibody $a$, much if the binding is lost (if it is the strongest site of escape, all binding is lost).
The variable $s$ represents how dramatically binding is lost for mutations at sites of escape that are not the strongest: larger values of $s$ means mutations at even moderate sites of escape reduce binding a lot.
In this paper we report calculations with $s = 2$, although the estimator has an option (\texttt{mutation\_escape\_strength}) to choose other values.
We then define the escape map after the mutations $\mathcal{M}$ as $\frac{1}{A}\sum_a x_{a,r} b_a\left(\mathcal{M}\right)$.
The estimator shows the escape map with no mutations in gray, and that after mutations in blue.

@@

\subsection{Implementation of interactive estimator}
The interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} is imlemented using Altair (\url{https://altair-viz.github.io/})~\citep{VanderPlas2018}, which is in turn built upon Vega-Lite~\citep{Satyanarayan2017}.

\subsection{Python module with batch-mode estimator}
A Python module that implements the calculations is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}, and has all the same options as the interactive estimator.
